Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 numerical.tex
restructured
appendix.tex

This setup is similar to that adopted
in \cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code.

@@

{\bf Growth rates are expected to diminish for both $\khat\to 0$
  and $\khat\to \infty$. This is because for $\khat\ll 1$ the radial
  wavelenth is large, but such disturbances are 
  stabilized by by epicyclic motions ($|\omega|\sim
  \kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
  there is no energy change when fluid elements are perturbed at fixed
  cylindrical radius whilst conserving its angular momentum (since we
  consider axisymmetric perturbations).  
}

@@

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in \S\ref{iso_poly}. For completeness we show the
corresponding eigenfunctions in vertical velocity in
Fig. \ref{lowfreq_eigenfuncvz}. The fundamental mode
consists of zero nodes in $\delta v_z$.

For $\khat\lesssim 10$, e.g. $\khat=1$ in 
Fig. \ref{lowfreq_eigenfunc}, the perturbation magnitude near the
boundaries begin to dominate. For even smaller $\khat\lesssim 1$ (not
shown) we find the perturbation is almost entirely concentrated at the
boundaries. Our radially-local approximation is not valid for 
these small wavenumber perturbations and we do not consider such modes
further.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvector_iso_kx1} 
  \includegraphics[width=\linewidth]{figures/eigenvector_iso_kx60}
  \caption{Eigenfunction $W$ of the fundamental VSI for perturbation
    wavenumber $\khat=1$ (top) and $\khat=20\pi$ (bottom). The real 
    (imaginary) parts of $W$ are plotted as solid (dotted) lines. The
    eigenfunction has been normalized such that $\imag(W)=0$ at the
    vertical boundaries and $\mathrm{max}|W|=1$. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental has the
smallest $|\sigma|$, for which the eigenvalue calculated from
Eq. \ref{sig2_iso} with $L=1$ and the numerically-obtained value are   
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surprisingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrals with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

@@

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides an
accurate prediction of the upper limit to the thermal relaxation 
timescale for  the fundamental mode. For smaller $\khat$,
$\beta<\beta_\mathrm{crit}$ is a sufficient condition for
instability. For larger $\khat$, we find $\nu$ does not reach zero but
rapidly decreases as $\beta$ approaches and increases beyond
$\beta_\mathrm{crit}$. In the latter case we find the perturbation is
concentrated at the vertical boundaries, which may play a role not
accounted by our analytical discussion.

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the vertical velocity
perturbation $\delta v_z$ for
$\beta=0.01$ and $\beta=0.1$. The latter value is close to the
critical value beyond which the fundamental VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Increasing $\beta$ 
restricts  the region in which $\delta v_z\sim $ constant closer to the midplane, with
oscillatory behavior emerging at the boundaries, where buoyancy first
becomes important. Finite thermal relaxation increases $|\delta v_z|$ near the
vertical boundaries relative to that about the midplane. However, 
perturbations near the disk surface may not be energetically important
in practice because the disk atmosphere contains little mass.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorvz_beta0d1} 
  \caption{Numerically-calculated fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta = 0.01$
    (top) and $\beta=0.1$ (bottom). The
    real (imaginary) part of the vertical velocity perturbation
    $\delta v_z$ is plotted as the solid (dotted)
    line. The eigenfunction has been normalized such that
    $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$.
    \label{relax_eigenW_num}}  
\end{figure}

>>>>

This setup is similar to that adopted
in \cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code. We also compare results with our 
analytical discussion of isothermal perturbations in Appendix
\ref{iso_discuss}.

@@

 Growth rates are expected to diminish for both $\khat\to 0$
  and $\khat\to \infty$. This is because for $\khat\ll 1$ the radial
  wavelenth is large, but such disturbances are 
  stabilized by by epicyclic motions ($|\omega|\sim
  \kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
  there is no energy change when fluid elements are perturbed at fixed
  cylindrical radius whilst conserving its angular momentum (since we
  consider axisymmetric perturbations).

@@

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in Appendix \ref{iso_poly}. For completeness we show the
corresponding eigenfunctions in vertical velocity in
Fig. \ref{lowfreq_eigenfuncvz}. The fundamental mode
consists of zero nodes in $\delta v_z$.

For $\khat\lesssim 10$, e.g. $\khat=1$ in 
Fig. \ref{lowfreq_eigenfunc}, the perturbation magnitude near the
boundaries begin to dominate. For even smaller $\khat\lesssim 1$ (not
shown) we find the perturbation is almost entirely concentrated at the
boundaries. However, our radially-local approximation is not valid for 
such small wavenumber perturbations.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvector_iso_kx1} 
  \includegraphics[width=\linewidth]{figures/eigenvector_iso_kx60}
  \caption{Eigenfunction $W$ of the fundamental VSI for perturbation
    wavenumber $\khat=1$ (top) and $\khat=20\pi$ (bottom). The real 
    (imaginary) parts of $W$ are plotted as solid (dotted) lines. % The
    % eigenfunction has been normalized such that $\imag(W)=0$ at the
    % vertical boundaries and $\mathrm{max}|W|=1$. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental mode has the
smallest $|\sigma|$, for which the eigenvalue calculated from
Eq. \ref{sig2_iso} with $L=1$ and the numerically-obtained value are   
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surprisingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrals with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

@@

For $5\lesssim\khat\lesssim30$, we find $\beta_\mathrm{crit}$ provides an
accurate prediction of the upper limit to the thermal relaxation 
timescale for  the fundamental mode. For smaller $\khat$,
$\beta<\beta_\mathrm{crit}$ is a sufficient condition for
instability. For larger $\khat$, we find $\nu$ does not reach zero but
rapidly decreases as $\beta$ approaches and increases beyond
$\beta_\mathrm{crit}$. In the latter case we find the perturbation
acquires large amplitudes near the vertical boundaries. Boundary
conditions may thus may play a role for these modes that is not
accounted by our analytical discussion.

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the vertical velocity
perturbation $\delta v_z$ for
$\beta=0.01$ and $\beta=0.1$. The latter value is close to
$\beta_\mathrm{crit}=0.125$. 
% the
% critical value beyond which the fundamental VSI is stabilized
% (\S\ref{iso_vsi_beta_crit}). 
Increasing $\beta$ 
restricts  the region in which $\delta v_z\sim $ constant closer to the midplane, with
oscillatory behavior emerging at the boundaries, where buoyancy first
becomes important. Finite thermal relaxation increases $|\delta v_z|$ near the
vertical boundaries relative to that about the midplane. However, 
perturbations near the disk surface may not be energetically important
in practice because the disk atmosphere contains little mass.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorvz_beta0d1} 
  \caption{Numerically-calculated fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta = 0.01$
    (top) and $\beta=0.1$ (bottom). The
    real (imaginary) part of the vertical velocity perturbation
    $\delta v_z$ is plotted as the solid (dotted)
    line. % The eigenfunction has been normalized such that
    % $\imag(\delta v_z)=0$ at the vertical boundaries, and
    % $\mathrm{max}|\delta v_z|=1$.
    \label{relax_eigenW_num}}  
\end{figure}

### 2015-03-04 paper.tex
appendix.tex

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include a thermal
  relaxation term that restores the disk  
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear calculations 
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI at tens of AU with
  a characteristic growth times of $\sim 30$ local orbital periods. 
\end{abstract}

@@

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually
assumed for typical disk models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the viability of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk become turbulent through the VSI, it must pass through the
linear instability first. Thus, understanding the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks in which the disk  
  temperature $T$ is restored to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying thermal timescale $\beta$. We find increasing $\beta$ even slightly away from 
  zero can dramatically reduce VSI growth rates. % , since this generally 
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. Our  linear calculations are consistent with previous hydrodynamic 
  simulations. We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital
  periods, and the dominant disturbance in the MMSN is the `corrugation' mode
  which casues the disk column to oscillate vertically. 
\end{abstract}

@@

From these studies it is clear that the VSI is generally sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually
assumed for typical disk models,  
% disk thermodynamics except
% in the special case of neutrally stratified disks. 
it is 
important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the relevance of
the VSI in realistic disks. % Numerical simulations of the VSI are
% challenging, however, because global disk simulations should be
% carried out while resolving radial scales much smaller than the local
% disk scale height.    
%On the other hand, 
Should a given disk become turbulent through the VSI, it must pass through the
linear instability first. Thus, understanding the thermodynamic dependence
of the linear VSI is essential. % Furthermore, linear
% stability calculations are essential to complement future numerical simulations.

### 2015-03-04 setup.tex
appendix.tex

We consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. %For $\Gamma=\gamma$ the disk is adiabatically stratified
% in the vertical 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index, and
define   
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $\epsilon$
is also a characteristic disk aspect-ratio.  In practice, we consider
thin disks with $\epsilon \ll 1$.

@@

\subsubsection{Vertically isothermal disks}
For $\Gamma=1$ the equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ are given by 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks which is the relevant case for irradiated
protoplanetary disks \citep{chiang97}.

\subsection{Vertical shear}\label{vshear_def}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot} and Eq. \ref{vertiso_eqm}, but is also 
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}
Evaluating this expression explicitly,
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz\Omega_k^2}{\Gamma
    r\left(1+z^2/r^2\right)^{3/2}}. \label{vertical_shear_ex} 
\end{align}
Note that there is a maximum vertical shear rate at $z=r/\sqrt{2}$,
which is expected to limit the growth rate of unstable modes.

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum. For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we consider the second
criterion. For our disk models,
\begin{align}
  &\frac{\p S}{\p r} = c_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = c_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z}. 
\end{align} 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\Omega_k^2$ but
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O(\epsilon^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
thermal relaxation, as shown below.

@@

In order to emphasize the radially-local approximation, we
introduce the local Cartesian co-ordinates $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$
using their expressions in the global disk given in \S\ref{setup}. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.

>>>>

We consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. %For $\Gamma=\gamma$ the disk is adiabatically stratified
% in the vertical 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index, and
define   
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $\epsilon$
is also a characteristic disk aspect-ratio. We are 
interested in thin disks with $\epsilon \ll 1$.%, as relevant for protoplanetary disks \citep{chiang97}.

@@

\subsubsection{Vertically isothermal disks}
For $\Gamma=1$ the equilibrium density $\rho(r,z)$ and rotation
$\Omega(r,z)$ are given by 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{\epsilon^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]}\\    
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}& =1+ (p+q)\epsilon^2(r) + q\left(1 -
    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
\end{align}
which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
vertically isothermal disks which is the relevant case for passively irradiated
protoplanetary disks \citep{chiang97}.

\subsection{Vertical shear}\label{vshear_def}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot} and Eq. \ref{vertiso_eqm}, but is also 
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}
Evaluating this expression explicitly,
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz\Omega_k^2}{\Gamma
    r\left(1+z^2/r^2\right)^{3/2}}. \label{vertical_shear_ex} 
\end{align}
Note that there is a maximum vertical shear rate at $z=r/\sqrt{2}$,
which is expected to limit the growth rate of instabilities associated with vertical shear.

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum. For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we consider the second
criterion. For our disk models,
\begin{align}
  &\frac{\p S}{\p r} = c_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = c_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z}. 
\end{align} 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\Omega_k^2$ but
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O(\epsilon^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
thermal relaxation, as shown below.

@@

In order to emphasize the radially-local approximation, we
introduce the local Cartesian co-ordinates $(x,y,z)$ corresponding to
the the local $(r,\phi,z)$ directions, respectively. Then writing
$\delta P /\rho \equiv W$ and $c_s^2\delta\rho/\rho\equiv Q$, the
linearized system of equations are 
\begin{align}
  \ii\sigma \frac{Q}{c_s^2}  &=  \ii k_x \delta v_x + \frac{d\dd
    v_z}{dz} + \frac{\p\ln{\rho}}{\p z}\delta v_z + \hat{g}_c
  \frac{\p\ln{\rho}}{\p r} \delta v_x,\label{lin_mass}\\
  \ii\sigma \delta v_x  &= \ii k_x W - 2\Omega\delta v_y -
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\frac{Q}{c_s^2}\label{lin_vx}\\
   \ii \sigma\delta v_y &= r\frac{\p \Omega}{\p z}\delta v_z +
  \frac{\kappa^2}{2\Omega}\delta v_x, \label{lin_vy}\\
   \ii\sigma\delta v_z &= \frac{dW}{dz} +
  \frac{\p\ln{\rho}}{\p z}\left(W-Q\right), \label{lin_vz}\\
  \ii \sigma W &= c_s^2\frac{\p\ln{\rho}}{\p z}\delta v_z +
  c_s^2\frac{\gamma}{\Gamma} \left(\ii k_x\delta v_x + \frac{d\delta
      v_z}{dz}\right) \notag\\
  &+ \frac{1}{t_c}\left(W - \frac{Q}{\Gamma}\right) +
  \hat{g}_c\frac{1}{\rho}\frac{\p P}{\p r}\delta v_x,\label{lin_energy}
  % -&\ii\sigma W + \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right) = -\ii\sigma\frac{\gamma}{\Gamma} Q +
  % c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma}-1\right)\delta v_z.\label{lin_energy}
\end{align}
where we have made the notation change $k_r\to k_x$. 
It is understood that the quantities $\rho,\,c_s^2,\,\Omega,\kappa,\, 
\p_z\Omega,\, \p_r\ln{\rho} $ and $ \p_r P$ are evaluated at $r=r_0$
using their expressions in the global disk given in \S\ref{setup}. Thus, they are functions of $z$ only. 
Eq. \ref{lin_mass}---\ref{lin_energy} will form the basis of our numerical
study, but for analytical discussion
we will make further simplifications.
